---
title: "Estimating Subgroup Treatment Effects with the Double Shrinker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Subgroup Treatment Effects with the Double Shrinker}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library( tidyverse )
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

Say we have two data sources on the effect of a treatment on coronary heart disease (`Outcome.CHD`):

- An **RCT** with clean identification but limited sample size.
- An **observational study (OS)** with many more units but potential confounding bias.

The double shrinker combines both sources to produce stratum-specific treatment effect estimates that are more precise than the RCT alone, while protecting against bias by shrinking toward the RCT when the two sources disagree.

This vignette walks through how to conduct such an analysis using the `doubleShrinker` package, using some fake data generated to mimic the actual (privacy-protected) data used in the paper associated with this work.

```{r load}
library(doubleShrinker)


load( here::here( "data/data.rct.RData" ) )
head( data.rct )
nrow( data.rct )


load( here::here( "data/data.os.RData" ) )
head( data.os )
nrow( data.os )

```

Note we are assuming we have a propensity score from some prior work we have done on the observational study.
The nice thing about this approach is we do not need to worry about where it came from.
If our observational study is well-specified, then the double shrinker will learn to trust it more; if it is misspecified, the double shrinker will learn to trust it less.


## Step 1: Define subgroups

We subgroup by two variables — `CVD` (pre-existing cardiovascular disease) crossed with `AGER` (age category) — giving us several strata to analyze jointly.

```{r subgroups}
subgroupVars <- c("CVD", "AGER")
outcome      <- "Outcome.CHD"

allData <- subgroupDefs_noSplit(
  subgroupVars = subgroupVars,
  data.os      = data.os,
  data.rct     = data.rct,
  outcome      = outcome,
  treatment    = "Test"
)

rctData <- allData$data.rct
obsData <- allData$data.os
k       <- length(allData$subgroupTaus.true)

cat("Number of strata:", k, "\n")
cat("Strata:\n")
print(allData$subgroupTaus.true)
```

The `subgroupTaus.true` vector gives treatment effect estimates for each stratum, computed from the full RCT population.

## Step 2: Estimate per-stratum effects

`getParamEstimates()` computes stratum-level treatment effect estimates and sampling variances from both data sources. The OS estimate uses inverse probability weighting (IPW) to adjust for confounding.

```{r estimates}
ed <- getParamEstimates(rctData, obsData, outcome, treatment = "Test")
print(ed)
```
Each row is a stratum. `rctEst`/`rctVar` are the RCT difference-in-means estimate and its variance; `obsEst`/`obsVar` are the IPW-adjusted OS estimate and its variance.
One could build this table by hand, using whatever methods for analysis one would like.

Notice that the OS variances are generally smaller (more data), but the RCT and OS estimates sometimes disagree — that disagreement, across strata, is what the double shrinker quantifies as `gammaSq` (bias variance).

We can look at a scatterplot of our estimates to see if there is agreement:

```{r, include=FALSE}
ed2 <- ed %>%
  mutate(
    a = 1.96 * sqrt(rctVar),   # horizontal radius
    b = 1.96 * sqrt(obsVar), # vertical radius
    id = row_number()

  )

ggplot(ed2, aes(rctEst, obsEst, color=as.factor(id))) +
    ggforce::geom_ellipse(
    aes(x0 = rctEst, y0 = obsEst, a = a, b = b, angle = 0,
        fill = as.factor(id) ),
    color = NA,
    alpha = 0.25
  ) +
  geom_point(size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_fixed() +
  theme_minimal() +
  theme( legend.position = "none" )
```
There is little agreement, save for the two highly precise estimates, suggesting lots of noise or lots of bias, or both.


## Step 3: Apply the double shrinker estimators

We apply all four hyperparameter estimation methods. We use `lowerBound = TRUE` to prevent near-zero hyperparameter estimates, which is recommended when the number of strata is small.

```{r double-shrinker}
est_mm1 <- eb.mm1(ed, k, lowerBound = TRUE)
est_mm2 <- eb.mm2(ed, k, lowerBound = TRUE)
est_mle <- eb.mle(ed, k, lowerBound = TRUE)
est_ure <- eb.ure(ed, k, lowerBound = TRUE)

# Make a tibble of results
ests <- tibble(
  strata = ed$stratum,
  rct_only   = ed$rctEst,
  obs_only   = ed$obsEst,
  mm1 = est_mm1,
  mm2 = est_mm2,
  mle = est_mle,
  ure = est_ure
)
```

## Step 4: Apply competitor estimators

For comparison, we also apply the precision-weighted combination and the Rosenman et al. (2023) kappa estimators.

```{r competitors}

ests <- ests %>%
  mutate( est_prec = precistion.wtd(ed, k),
          kappa1 = kappa.1(ed, k),
          kappa2 = kappa.2(ed, k),
          delta1 = delta.1(ed, k),
          delta2 = delta.2(ed, k)
  )
```

We finally print out our results:

```{r}
knitr::kable( ests, digits=3 )
```


A few things to notice:

- The **RCT-only** estimates are unbiased but noisy; they vary quite a bit across strata relative to the true effects.
- The **OS-only** estimates are more stable but may be systematically off due to confounding.
- The **double shrinker** variants (prefixed `ds_`) generally sit between the two, pulled toward the OS when its estimates are consistent with the RCT, and toward the RCT when they disagree.
- The **precision-weighted** estimator also blends both sources, but does not account for OS bias — it will over-weight the OS whenever `obsVar` is small, regardless of disagreement.

## Step 6: Empirical Bayes confidence intervals

The `eb.mm1` estimator can return the components needed to construct valid confidence intervals that account for the shrinkage.

```{r ci}
eb_comps <- eb.mm1(ed, k, lowerBound = TRUE, returnCIComps = TRUE)

half_widths <- ebCiFunc(eb_comps, ed, alpha = 0.05)

ci <- data.frame(
  stratum  = ed$stratum,
  true_tau = allData$subgroupTaus.true,
  estimate = eb_comps$shrinker,
  lower    = eb_comps$shrinker - half_widths,
  upper    = eb_comps$shrinker + half_widths
)

ci %>%
  knitr::kable( digits = 3 )

```

These are empirical Bayes intervals — they use a critical value adjusted for the shrinkage bias, so they have better coverage properties than naive normal intervals around the shrunken point estimate.

## Summary

In this analysis the double shrinker combines a small RCT and a larger observational study to produce more precise stratum-level treatment effect estimates. The four hyperparameter estimation strategies (MM1, MM2, MLE, URE) all target the same model and tend to give similar estimates.
The `lowerBound = TRUE` option is advisable with small numbers of strata to prevent degenerately small hyperparameter estimates.


